\chapter{List of examples of affine schemes}
To cement in the previous chapter,
we now give an enormous list of examples.
Each example gets its own section,
rather than having page-long orange boxes.

\section{Example: $\Spec k$, a single point}
This one is easy: for any field $k$,
$X = \Spec k$ has a single point,
corresponding to the only proper ideal $(0)$.
There is only way to put a topology on it.

As for the sheaf,
\[ \OO_X(X) = \OO_{X,(0)} = k. \]
So the space is remembering what field it wants to be over.
If we are complex analysists,
the set of functions on a single point is $\CC$;
if we are number theoerists,
maybe the set of functions on a single point is $\QQ$.

%%fakechapter One-dimensional
\section{$\Spec \CC[x]$, a one-dimensional line}
The scheme $X = \Spec \CC[x]$ is our beloved one-dimensional line.
It consists of two types of points:
\begin{itemize}
	\ii The closed points $(x-a)$, corresponding to each
	complex number $a \in \CC$, and
	\ii The \emph{generic} point $(0)$.
\end{itemize}
As for the Zariski topology, every open set contains $(0)$,
which captures the idea it is close to everywhere:
no matter where you stand, you can still here the buzzing of the fly!
True to the irreducibility of this space,
the open sets are huge:
the proper \emph{closed sets} consist of finitely many closed points.

Here is a picture:
for lack of better place to put it,
the point $(0)$ is floating around just above the line in red.
\begin{center}
	\begin{asy}
		size(8cm);
		pair A = (-8,0); pair B = (8,0);
		draw(A--B, Arrows);
		dot("$(x)$", (0,0), dir(-90), blue);
		dot("$(x-3)$", (3,0), dir(-90), blue);
		dot("$(x+i)$", (-4,0), dir(-90), blue);
		dot("$(0)$", (0,1), dir(90), red);
		draw( (-6,0)..(-3,0.5)..(-1,1)--(1,1)--(3,0.5)--(6,0), dotted+red );
		label("$\operatorname{Spec} \mathbb C[x]$", A, dir(-90));
	\end{asy}
\end{center}

The notion of ``value at $\kp$'' works as expected.
For example, $f = x^2+5$ is a global section of $\CC[x]$.
If we evaluate it at $\kp = x-3$,
we find $f(\kp) = f \pmod \kp = x^2+5 \pmod{x-3} = 14 \pmod{x-3}$.
As \[ \CC[x] / \kp \cong \CC \qquad \text{by $x \mapsto 3$} \]
we see we are just plugging $x=3$.

Of course, the stalk at $(x-3)$ carries more information.
In this case it is $\CC[x]_{(x-3)}$.
Which means that if we stand near the point $(x-3)$,
rational functions are all fine as long as no $x-3$
appears in the denominator.
So, $\frac{x^2+8}{(x-1)(x-5)}$ is a fine example of a germ near $x=3$.

Things get more interesting if we
consider the generic point $\eta = (0)$.
What happens if we evaluate the global section
$f = x^2+5$ at $\eta$?
Well, we just get $f(\eta) = x^2+5$ ---
taking modulo $0$ doesn't do much.
Fitting, it means that if you want to be able to evaluate
a polynomial $f$ at a general complex number,
you actually just need the whole polynomial.

What is the stalk $\OO_{X, \eta}$?
Well, it should be $\CC[x]_{(0)} = \CC(x)$,
which is the set of \emph{rational} functions.
And that's what you expect.
For example, $\frac{x^2+8}{(x-1)(x-5)}$
certainly describes a rational function on ``most'' complex numbers.

\section{$\Spec \RR[x]$, a one-dimensional line
with complex conjugates glued (no fear nullstellensatz)}

Despite appearances, this actually
looks almost exactly like $\Spec \CC[x]$,
even more than you expect.
The main thing to keep in mind is that now $(x^2+1)$
is a point, which you can loosely think of as $\pm i$.
So it almost didn't matter that $\RR$ is not algebraically closed;
the $\CC$ is showing through anyways.
But this time, because we only consider
real coefficient polynomials,
we do not distinguish between ``conjugate'' $+i$ and $-i$.
Put another way, we have folded $a+bi$ and $a-bi$ into a single point:
$(x+i)$ and $(x-i)$ merge to form $x^2+1$.

To be explicit, there are three types of points:
\begin{itemize}
	\ii $(x-a)$ for each real number $a$
	\ii $(x^2-ax+b)$ if $a^2 < 4b$, and
	\ii the generic point $(0)$, again.
\end{itemize}
The ideals $(x-a)$ and $(x^2-ax+b)$
are each closed points:
the quotients with $\RR[x]$ are both fields
($\RR$ and $\CC$, respectively).


We have been drawing $\Spec \CC[x]$ as a one-dimensional line,
so $\Spec \RR[x]$ will be drawn the same way.
\begin{center}
	\begin{asy}
		size(8cm);
		pair A = (-8,0); pair B = (8,0);
		draw(A--B, Arrows);
		dot("$(x)$", (0,0), dir(-90), blue);
		dot("$(x-3)$", (3,0), dir(-90), blue);
		dot("$(x^2+1)$", (-4,0), dir(-90), blue);
		dot("$(0)$", (0,1), dir(90), red);
		draw( (-6,0)..(-3,0.5)..(-1,1)--(1,1)--(3,0.5)--(6,0), dotted+red );
		label("$\operatorname{Spec} \mathbb R[x]$", A, dir(-90));
	\end{asy}
\end{center}

One nice thing about this is that the nullstellensatz
is less scary than it was with classical varieties.
The short version is that the function $x^2+1$
vanishes at a point of $\Spec \RR[x]$,
namely $(x^2+1)$ itself!

You might remember a long time ago we made a big fuss about
the weak nullstellensatz, for example in
\Cref{pr:complex_variety_nonempty}:
if $I$ was a proper ideal in $\CC[x_1, \dots, x_n]$
there was \emph{some} point $(a_1, \dots, a_n) \in \CC^n$
such that $f(a_1, \dots, a_n) = 0$ for all $f \in I$.
With schemes, it doesn't matter anymore:
if $I$ is a proper ideal of a ring $A$,
then some maximal ideal contains it,
and so $\VV(I)$ is nonempty in $\Spec A$.

\section{$\Spec k[x]$, over any ground field}
In general, if $\ol k$ is the algebraic closure of $k$,
then $\Spec k[x]$ looks like $\Spec \ol k[x]$
with all the Galois conjugates glued together.
So we will almost never need ``algebraically closed''
hypotheses anymore: we're working with polynomial ideals,
so all the elements are implicitly there, anyways.

\section{$\Spec \ZZ$, a one-dimensional scheme}
The great thing about $\Spec \ZZ$ is that
it basically looks like $\Spec k[x]$, too,
being a one-dimensional scheme.
It has two types of prime ideals:
\begin{itemize}
	\ii $(p)$, for every rational prime $p$,
	\ii and the generic point $(0)$.
\end{itemize}
So the picture almost does not change.
\begin{center}
	\begin{asy}
		size(8cm);
		pair A = (-8,0); pair B = (8,0);
		draw(A--B, Arrows);
		dot("$(7)$", (0,0), dir(-90), blue);
		dot("$(3)$", (3,0), dir(-90), blue);
		dot("$(19)$", (-4,0), dir(-90), blue);
		dot("$(0)$", (0,1), dir(90), red);
		draw( (-6,0)..(-3,0.5)..(-1,1)--(1,1)--(3,0.5)--(6,0), dotted+red );
		label("$\operatorname{Spec} \mathbb R[x]$", A, dir(-90));
	\end{asy}
\end{center}
This time $\eta = (0)$ has stalk $\ZZ_{(0)} = \QQ$,
so a ``rational function'' is literally a rational number!
Thus, $\frac{20}{19}$ is a function
with a double root at $(2)$, a root at $(5)$,
and a simple pole at $(19)$.
If we evaluate it at $\kp = (3)$, we get $2 \pmod 3$.

The stalk $X_{(3)}$ is the rational numbers
that don't have a pole at $3$.

%%fakechapter 0D from 1D
\section{$\Spec k[x] / (x^2-x)$, two points}
If we were working with affine varieties,
you would already know what the answer is:
$x^2-x = 0$ has solutions $x=0$ and $x=1$,
so this should be a scheme with two points.

To see this come true, we use \Cref{prop:prime_quotient}:
the points of $\Spec k[x]/(x^2-x)$
should correspond to prime ideals of $k[x]$
contained in $(x^2-x)$.
As $k[x]$ is a PID, there are only two, $(x-1)$ and $(x)$.
They are each maximal,
since their quotient with $k[x]$ is a field (namely $k$),
so as promised $\Spec k[x] / (x^2-x)$ has just two closed points.

Each point has a stalk above it isomorphic to $k$.
A section on the whole space $X$ is just a choice
of two values, one at $(x)$ and one at $(x-1)$.
\begin{center}
\begin{asy}
	size(3cm);
	draw("$k$", (-2,0)--(-2,5), dir(180), red);
	draw("$k$", (2,0)--(2,5), red);
	dot("$(x)$", (-2,0), dir(-90), blue);
	dot("$(x-1)$", (2,0), dir(-90), blue);
	label("$\operatorname{Spec} k[x]/(x^2-x)$", (0,5), 2*dir(90));
\end{asy}
\end{center}
So actually, this is a geometric way of thinking about the
ring-theoretic fact that
\[ k[x] / \left( x^2-x \right) \cong k \times k
	\quad\text{by } f \mapsto \left( f(0), f(1) \right).  \]

Also, this is the first example of a reducible space in this chapter:
in fact $X$ is even disconnected.
Accordingly there is no generic point floating around:
as the space is discrete, all points are closed.

\section{$\Spec k[x]/(x^2)$, the double point}
We can now elaborate on the ``double point'' scheme
\[ X_2 = \Spec k[x] / (x^2) \]
since it is such an important motivating example.
How it does differ from the ``one-point'' scheme
$X_1 = \Spec k[x] / (x) = \Spec k$?
Both $X_2$ and $X_1$ have exactly one point,
and so obviously the topologies are the same too.

The difference is that the stalk
(equivalently, the section, since we have only one point)
is larger:
\[ \OO_{X_2}(X_2) = \OO_{X_2, (x)} = k[x]/(x^2). \]
So to specify a function on a double point,
you need to specify two parameters, not just one:
if we take a polynomial
\[ f = a_0 + a_1 x + \dots \in k[x] \]
then evaluating it at the double point
will remember both $a_0$ and the ``first derivative'' say.

\section{$\Spec k[x]/(x^3-x)$, a double point and a single point}
There is no problem putting the previous two examples side by side:
the scheme $X = \Spec k[x] / (x^3-x)$
consists of a double point next to a single point.
Note that the stalks are different:
the one above the double point is larger.
\begin{center}
\begin{asy}
	size(5cm);
	draw("$k[x] / (x^2)$", (-2,0)--(-2,7), dir(180), red);
	draw("$k$", (2,0)--(2,5), red);
	dot("$(x)$", (-2,0), dir(-90), blue+5);
	dot("$(x-1)$", (2,0), dir(-90), blue);
	label("$\operatorname{Spec} k[x]/(x^3-x)$", (0,7), 2*dir(90));
\end{asy}
\end{center}
This time, we implicitly have the ring isomorphism
\[ k[x] / (x^3-x) \cong k[x] / (x^2) \times k \]
by $f \mapsto \left( f(0) + f'(0) x, f(1) \right)$.
The derivative is meant formally here!

\section{$\Spec \Zc{60}$, a scheme with three points}
We've being seeing geometric examples of ring products coming up,
but actually the Chinese remainder theorem you are used to
with integers is no different.
(This example $X = \Spec \Zc{60}$
is taken from \cite[\S4.4.11]{ref:vakil}.)

By \Cref{prop:prime_quotient},
the prime ideals of $\Zc{60}$ are $(2)$, $(3)$, $(5)$.
But you can think of this also as coming out of $\Spec \ZZ$:
as $60$ was a function with a double root at $(2)$,
and single roots at $(3)$ and $(5)$.
\begin{center}
\begin{asy}
	size(6cm);
	draw("$\mathbb Z / 4 \mathbb Z$", (-2,0)--(-2,7), dir(180), red);
	draw("$\mathbb Z / 3 \mathbb Z$", (0,0)--(0,5), red);
	draw("$\mathbb Z / 5 \mathbb Z$", (2,0)--(2,6), red);
	dot("$(2)$", (-2,0), dir(-90), blue+5);
	dot("$(3)$", ( 0,0), dir(-90), blue);
	dot("$(5)$", ( 2,0), dir(-90), blue);
	label("$\operatorname{Spec} \mathbb Z / 60 \mathbb Z$", (0,7), 2*dir(90));
\end{asy}
\end{center}
Actually, although I have been claiming the ring isomorphisms,
the sheaves really actually give us a full proof.
Let me phrase it in terms of global sections:
\begin{align*}
	\Zc{60} &= \OO_X(X) \\
	&= \OO_X( \{(2)\} ) \times \OO_X( \{(3)\} ) \times \OO_X( \{(5)\} ) \\
	&= \OO_{X,(2)} \times \OO_{X,(3)} \times \OO_{X,(5)} \\
	&= \Zc4 \times \Zc3 \times \Zc5.
\end{align*}
So the theorem that $\OO_X(X) = A$ for $X = \Spec A$
is doing the ``work'' here;
the sheaf axioms then give us the Chinese remainder theorem from here.

%%fakechapter Two-dimensional
\section{$\Spec k[x,y]$, the two-dimensional plane}
\section{$\Spec \ZZ[x]$, a strange two-dimensional scheme}

%%fakechapter 1D cut out from 2D
\section{$\Spec k[x,y]/(y-x^2)$, the parabola}
\section{$\Spec k[x,y]/(xy)$, two axes}
\section{$\Spec k[x] \times k[y]$, two ``parallel'' lines}

%%fakechapter 1D with holes
\section{$\Spec k[x]_x$, the punctured line (or hyperbola)}
\section{$\Spec \ZZ_{55}$, deleting $5$ and $11$}
\section{$\Spec (k[x,y]/(xy))_{x+y}$, two axes with the origin deleted}
%%fakechapter 0D --- stalks above 1D
\section{$\Spec k[x]_{(x)}$, the stalk above the origin}
\section{$\Spec k[x]_{(0)}$, the stalk above the generic point}
\section{$\Spec \ZZ_{(5)}$, the stalk above $(5)$}
\section{$\Spec \ZZ_{(0)} = \Spec \QQ$}
\section{$\Spec k[x,y]_{(x,y)}$, the stalk above the origin}
\section{$\Spec k[x,y]_{(y-x^2)}$, the stalk above the parabola}
\section{$\Spec k[x,y]_{(0)}$, the stalk above the generic point}

%%fakechapter Problems
\section{\problemhead}

\begin{dproblem}
	[Chinese remainder theorem]
	Consider $X = \Spec \Zc{60}$, which as a topological space has three points.
	By considering $\OO_X(X)$ prove the Chinese theorem
	\[ \Zc{60} \cong \Zc{4} \times \Zc{3} \times \Zc{5}. \]
	\begin{hint}
		Appeal to \Cref{prob:finite_sheaf}.
	\end{hint}
\end{dproblem}

