\documentclass[11pt]{scrreprt}
\input{../../tex/preamble}
\def\asydir{}
\addbibresource{../../references.bib}
\renewcommand{\gim}{*}
\renewcommand{\yod}{**}
\renewcommand{\kurumi}{***}

\begin{document}
\title{Algebraic Geometry I: Varieties}
\maketitle

\tableofcontents

\chapter{Quasi-projective varieties and regular maps}
\section{Quasi-projective varieties}
Here is yet another type of baby ringed space,
which will turn out to generalize both affine and projective varieties.
\begin{definition}
	A \vocab{quasi-projective variety} is
	an open set $X$ of a projective variety $V$.
	It is a baby ringed space $(X, \OO_X)$ too,
	because for any open set $U \subseteq X$
	we simply define $\OO_X(U) = \OO_V(U)$.
\end{definition}
Obviously, every projective variety is quasi-projective.
But now, we can interpret every affine variety $V$
as a quasi-projective variety by noticing that
\[ V \subseteq \Aff^n \to U_0 \subset \CP^n \]
where $U_0$ is the standard affine chart, isomorphic to $\Aff^n$.

\section{Maps of baby ringed spaces}
In order to make this precise I have to define
what the arrow $\Aff^n \to U_0$ means,
because they are different types of objects;
the former is an affine variety and the latter is a quasi-projective variety.
So, let me actually define a map between \emph{any} baby ringed spaces:
\begin{definition}
	Let $(X, \OO_X)$ and $(Y, \OO_Y)$
	be baby ringed spaces.
	A continuous map of topological spaces $f: X \to Y$
	is a \vocab{regular map} if the following condition holds:
	if $U$ is an open subset of $Y$.
	and $\phi \in \OO_Y(U)$ is regular, then the composition
	\begin{diagram}
		f\pre(U) &\rTo^f & U & \rTo^\phi & \CC
	\end{diagram}
	is a regular function in $\OO_X(f\pre(U))$.

	Two baby ringed spaces are \vocab{isomorphic}
	if there are mutually inverse regular maps between them.
\end{definition}
This is great.
Before this definition, we had three different types of varieties:
\begin{itemize}
	\ii Affine varieties,
	\ii Projective varieties, and
	\ii Quasi-projective varieties.
\end{itemize}
But all three are baby ringed spaces, now they can
talk to each other, and even be isomorphic.
In particular, let's now finally prove:
\begin{proposition}
	[Projective space is locally affine]
	The standard affine chart $U_0$ of $\CP^n$
	is isomorphic to $\Aff^n$.
\end{proposition}
\begin{proof}
	\todo{prove this}
\end{proof}

\section\problemhead
affine projective single point
\end{document}

\section{(Aside) Which rings are coordinate rings?}
As an aside, we might ask: which rings are the coordinate ring of some affine variety?
There are some obvious requirements.
\begin{itemize}
	\ii Such ring must be $\CC$-algebras, of course.
	(A $\CC$-algebra is a ring that contains a copy of $\CC$ in it.)
	\ii Such a ring must be finitely generated, since $\CC[V]$
	is generated by $x_1$, \dots, $x_n$.
\end{itemize}
There is a third more subtle condition.
Define an element $r \in R$ to be \vocab{nilpotent} if $r^N = 0$ for some $r \in R$.
For example, $x$ is nilpotent in $\CC[x]/(x^{2015})$.
Call a ring \vocab{reduced} if it has no nonzero nilpotent elements.

\begin{exercise}
	Let $I$ be an ideal of a ring $R$.
	Show that $I$ is radical if and only if $R/I$ is reduced.
\end{exercise}
\begin{remark}
	For a ring $R$, this completes the following table:
	\begin{center}
	\begin{tabular}[h]{l|l}
		Ideal $I$ & Quotient $R/I$ \\ \hline
		maximal & field \\
		prime & integral domain \\
		semiprime & reduced
	\end{tabular}
	\end{center}
	In particular, $(0)$ is maximal, prime, or radical
	if and only if $R/(0) \cong R$ is a field, integral domain,
	or reduced, respectively.
\end{remark}

Thus, the third requirement is that a coordinate ring must be reduced.
Conveniently, it turns out that these three conditions are sufficient!
\begin{theorem}[Coordinate rings are finitely generated reduced $\CC$-algebras]
	Every finitely generated reduced $\CC$-algebra is the coordinate
	ring of some complex affine variety.
\end{theorem}
\begin{proof}
	Let $R$ be this map.
	Because it's finitely generated by some $r_1$, $r_2$, \dots, $r_n$,
	there is some surjective map
	\[ \CC[x_1, \dots, x_n] \surjto R \]
	via $x_i \mapsto r_i$.
	Let $I$ be the kernel of this map.
	Since $R$ is reduced, $I$ is radical.
	Then $R \cong \CC[V]$, where $V = \Vp(I)$.
\end{proof}
