\chapter{A Tiny Dose of Category Theory}
I can't possibly hope to do category theory any justice in this chapter;
thus I'll just give a very high-level overview of how many of the concepts we've
encountered so far can be re-cast into categorical terms.
So I'll say what a category, give some examples, then talk about a few things that
categories can do.
For my examples, I'll be drawing from all the previous chapters;
feel free to skip over the examples corresponding to things you haven't seen.

If you're interested in category theory (like I was!), perhaps in
what surprising results are true for general categories, I strongly recommend \cite{ref:msci}.

\section{Motivation: Isomorphisms}
From earlier chapters let's recall the definition of an \emph{isomorphism} of two objects.
We had that
\begin{itemize}
	\ii Two groups $G$ and $H$ were isomorphic if there was a bijective homomorphism:
	equivalently, we wanted homomorphisms $\phi : G \to H$ and $\psi : H \to G$
	which were mutual inverses, meaning $\phi \circ \psi = \id_H$ and $\psi \circ \phi = \id_G$.
	\ii Two metric (or topological) spaces $X$ and $Y$ are isomorphic
	if there is a continuous bijection $f : X \to Y$ such that $f\inv$ is also continuous.
	\ii Two vector spaces $V$ and $W$ are isomorphic if there is a bijection $T : V \to W$
	which is a linear map.
	Again, this can be re-cast as saying that $T$ and $T\inv$ are continuous maps.
	\ii Two rings $R$ and $S$ are isomorphic if there is a bijective ring homomorphism $\phi$;
	again, we can recast this as two mutually inverse ring homomorphisms.
\end{itemize}

In each case we have some collections of objects and some maps,
and the isomorphisms can be viewed as just maps.
Let's use this to motivate the definition of a general \emph{category}.

\section{Categories, and Examples Thereof}
\prototype{$\catname{Gp}$ is possibly the most natural example.}
\begin{definition}
	A \vocab{category} $\AA$ consists of the following.
	\begin{itemize}
		\ii A class of \vocab{objects}, denoted $\obj(\AA)$.
		\ii For any two objects $A_1, A_2 \in \obj(\AA)$, 
		a class of \vocab{arrows} (also called \vocab{morphisms} or \vocab{maps}) between them.
		We'll denote the set of these arrows by $\Hom_\AA(A_1, A_2)$.
		\ii For any $A_1, A_2, A_3 \in \obj(A)$,
		if $f : A_1 \to A_2$ is an arrow and $g : A_2 \to A_3$ is an arrow, we can compose
		these arrows to get an arrow $g \circ f : A_1 \to A_3$.

		We can represent this in a \vocab{commutative diagram}
		\begin{diagram}
			A_1 & \rTo^f & A_2 \\
			& \rdDashed^{f \circ g} & \dTo_g \\
			&& A_3
		\end{diagram}

		The composition operation $\circ$ is part of $\AA$;
		it must be associative.
		
		\ii Finally, every object $A \in \obj(\AA)$ has a special \vocab{identity arrow} $\id_A$;
		you can guess what it does.\footnote{To be painfully explicit: if $f : A' \to A$ is an arrow then $\id_A \circ f = f$;
		similarly, if $g : A \to A'$ is an arrow then $g \circ \id_A = g$.}
	\end{itemize}
\end{definition}
\begin{remark}
	You can think of ``class'' as just ``set''.
	The reason we can't use the word ``set'' is because of some paradoxical issues with
	collections which are too large;
	Cantor's Paradox says there is no set of all sets.
	So referring to these by ``class'' is a way of sidestepping this issues.
\end{remark}
\begin{abuse}
	From now on, by $A \in \AA$ we'll mean $A \in \obj(\AA)$.
\end{abuse}

Let's formalize the motivation we began with.
\begin{example}
	[Basic Examples of Categories]
	\listhack
	\label{example:basic_categories}
	\begin{enumerate}[(a)]
		\ii There is a category of groups $\catname{Gp}$. The data is
		\begin{itemize}
			\ii The objects of $\catname{Gp}$ are the groups.
			\ii The arrows of $\catname{Gp}$ are the homomorphisms between these groups.
			\ii The composition $\circ$ in $\catname{Gp}$ is function composition.
		\end{itemize}
		\ii In the same way we can conceive a category $\catname{Ring}$ of rings.
		\ii Similarly, there is a category $\catname{Top}$ of topological spaces,
		whose arrows are the continuous maps.
		\ii There is a category $\catname{Top}_\ast$ of topological spaces with a \emph{distinguished basepoint},
		that is a pair $(X, x_0)$ where $x_0 \in X$.
		Arrows are continuous maps $f : X \to Y$ with $f(x_0) = y_0$.
		\ii Similarly, there is a category $\catname{Vect}_k$ of vector spaces over a field $k$,
		whose arrows are the linear maps.
		There is even a category $\catname{FDVect}_k$ of \emph{finite-dimensional} vector spaces.
		\ii We have a category $\catname{Set}$ of sets,
		where the arrows are \emph{any} maps.
	\end{enumerate}
\end{example}
And of course, we can now define what an isomorphism is!
\begin{definition}
	An arrow $A_1 \taking{f} A_2$ is an \vocab{isomorphism}
	if there exists $A_2 \taking{g} A_1$ such that $f \circ g = \id_{A_1}$
	and $g \circ f = \id_{A_2}$.
	In that case we say $A_1$ and $A_2$ are \vocab{isomorphic}, hence $A_1 \cong A_2$.
\end{definition}
\begin{remark}
	Note that in $\catname{Set}$, $X \cong Y
	\iff \left\lvert X \right\rvert = \left\lvert Y \right\rvert$.
\end{remark}
\begin{ques}
	Check that every object in a category is isomorphic to itself.
	(This is offensively easy.)
\end{ques}
More importantly, this definition should strike you as a little impressive.
We're able to define whether two groups (rings, spaces, etc.) are isomorphic
solely by the functions between the objects.
Indeed, one of the key themes in category theory (and even algebra) is that
\begin{moral}
	One can learn about objects by the functions betweens them.
	Category theory takes this to the extreme by \emph{only} looking at arrows,
	and ignoring what the objects themselves are.
\end{moral}

But there are some trickier interesting examples of categories.
\begin{example}
	[Posets are Categories]
	Let $\mathcal P$ be a partially ordered set.
	We can construct a category $P$ for it as follows:
	\begin{itemize}
		\ii The objects of $P$ are going to be the elements of $\mathcal P$.
		\ii The arrows of $P$ are defined as follows:
		\begin{itemize}
			\ii For every object $p \in P$, we add an identity arrow $\id_p$, and
			\ii For any pair of distinct objects $p \le q$, we add a single arrow $p \to q$.
		\end{itemize}
		There are no other arrows.
		\ii There's only one way to do the composition. What is it?
	\end{itemize}
\end{example}
For example, for the poset $\mathcal P$ on four objects $\{a,b,c,d\}$ with $a \le b$ and $a \le c \le d$, we get:
\begin{center}
\begin{tikzpicture}[scale=3.5]
	\SetVertexMath
	\Vertices{square}{d,c,a,b}
	\Edge[style={->}, label={$a \le b$}](a)(b)
	\Edge[style={->}, label={$a \le c$}](a)(c)
	\Edge[style={->}, label={$a \le d$}](a)(d)
	\Edge[style={->}, label={$c \le d$}](c)(d)
	\Loop[dist=8, dir=NO, label={$\id_a$}, labelstyle={above=1pt}](a)
	\Loop[dist=8, dir=WE, label={$\id_b$}, labelstyle={left=1pt}](b)
	\Loop[dist=8, dir=EA, label={$\id_c$}, labelstyle={right=1pt}](c)
	\Loop[dist=8, dir=WE, label={$\id_d$}, labelstyle={left=1pt}](d)
\end{tikzpicture}
\end{center}

This illustrates the point that
\begin{moral}
	The arrows of a category can be totally different from functions.
\end{moral}
In fact, in a way we'll make precise later, the term ``concrete category'' refers
to one where the arrows really are ``structure-preserving maps between sets'',
like $\catname{Gp}$, $\catname{Top}$, $\catname{Ring}$, \dots.

\begin{ques}
	Check that no two distinct objects of a poset are isomorphic.
\end{ques}

Here's a second quite important example of a non-concrete category.
\begin{example}
	[Important: Groups Are One-Object Categories]
	A group $G$ can be interpreted as a category $\mathcal G$ with one object $\ast$,
	all of whose arrows are isomorphisms.

	\begin{center}
	\begin{tikzpicture}[scale=5.5]
		\Vertex[x=0,y=0,L={$\ast$}]{a}
		\Loop[dist=8, dir=NO, label={$1 = \id_a$}, labelstyle={above=1pt}](a)
		\Loop[dist=7, dir=WE, label={$g_2$}, labelstyle={left=1pt}](a)
		\Loop[dist=9, dir=SO, label={$g_3$}, labelstyle={below=1pt}](a)
		\Loop[dist=8, dir=EA, label={$g_4$}, labelstyle={right=1pt}](a)
	\end{tikzpicture}
	\end{center}

	As \cite{ref:msci} says:

	\begin{quote}
	The first time you meet the idea that a group is a kind of category,
	it's tempting to dismiss it as a coincidence or a trick.
	It's not: there's real content.
	To see this, suppose your education had been shuffled and you took a course
	on category theory before ever learning what a group was.
	Someone comes to you and says: 

	``There are these structures called `groups', and the idea is this:
	a group is what you get when you collect together all the symmetries
	of a given thing.''

	``What do you mean by a `symmetry'?'' you ask.

	``Well, a symmetry of an object $X$ is a way of transforming $X$ or mapping
	$X$ into itself, in an invertible way.''

	``Oh,'' you reply, ``that's a special case of an idea I've met before.
	A category is the structure formed by \emph{lots} of objects and mappings
	between them -- not necessarily invertible. A group's just the very special case
	where you've only got one object, and all the maps happen to be invertible.''
	\end{quote}
\end{example}

\begin{exercise}
	Verify the above!
	That is, show that the data of a one-object category with all isomorphisms
	is the same of the data of a group.
\end{exercise}

Finally, here are some examples of categories you can make from other categories.
\begin{example}
	[Deriving Categories]
	\listhack
	\begin{enumerate}[(a)]
		\ii Given a category $\AA$, we can construct the \vocab{opposite category}
		$\AA^{\text{op}}$, which is the same as $\AA$ but with all arrows reversed.
		\ii Given a category $\AA$ and $\BB$, we can construct the \vocab{product category} $\AA \times \BB$
		as follows: the objects are pairs $(A, B)$ for $A \in \AA$ and $B \in \BB$,
		and the arrows from $(A_1, B_1)$ to $(A_2, B_2)$
		are pairs \[ \left( A_1 \taking{f} A_2, B_1 \taking{g} B_2 \right). \]
		What do you think the composition and identities are?
	\end{enumerate}
\end{example}

\section{Special Objects in Categories}
\prototype{$\catname{Set}$ has initial object $\varnothing$ and final object $\{\ast\}$. An element of $S$ corresponds to a map $\{\ast\} \to S$.}
Certain objects in categories have special properties.
Here are a couple examples.
\begin{example}
	[Initial Object]
	An \vocab{initial object} of $\AA$ is an object
	$A_{\text{init}} \in \AA$ such that for any $A \in \AA$ (possibly $A = A_{\text{init}}$),
	there is exactly one arrow from $A$ to $A_{\text{init}}$.
	For example,
	\begin{enumerate}[(a)]
		\ii The initial object of $\catname{Set}$ is the empty set $\varnothing$.
		\ii The initial object of $\catname{Gp}$ is the trivial group $\{1\}$.
		\ii The initial object of $\catname{Ring}$ is the ring $\ZZ$
		(recall that ring homomorphisms $R \to S$ map $1_R$ to $1_S$).
		\ii The initial object of $\catname{Top}$ is the empty space.
		\ii The initial object of a partially ordered set is its smallest element, if one exists.
	\end{enumerate}
\end{example}

We justify the use of ``the'' as follows.
\begin{exercise}
	In \emph{any} category,
	show that if an initial object exists,
	it's unique up to isomorphism.
\end{exercise}
\begin{remark}
	In mathematics, we really never know nor care if two objects are actually equal
	or whether they are isomorphic.
	For example, there are many competing ways to define $\RR$,
	but we still just refer to it as ``the'' real numbers.

	Thus when we define categorical notions, we would like to check they are
	unique up to isomorphism.
	This is really clean in the language of categories, and definitions
	often cause objects to be unique up to isomorphism for elegant reasons like the above.
\end{remark}

One can take the ``dual'' notion, a terminal object.
\begin{example}
	[Terminal Object]
	An \vocab{terminal object} of $\AA$ is an object
	$A_{\text{final}} \in \AA$ such that for any $A \in \AA$ (possibly $A = A_{\text{final}}$),
	there is exactly one arrow from $A$ to $A_{\text{final}}$.
	For example,
	\begin{enumerate}[(a)]
		\ii The terminal object of $\catname{Set}$ is the singleton set $\{\ast\}$.
		(There are many singleton sets, of course, but \emph{as sets} they are all isomorphic!)
		\ii The terminal object of $\catname{Gp}$ is the trivial group $\{1\}$.
		\ii The terminal object of $\catname{Ring}$ is the zero ring (a ring with element $0=1$).
		(recall that ring homomorphisms $R \to S$ map $1_R$ to $1_S$).
		\ii The terminal object of $\catname{Top}$ is the single-point space.
		\ii The terminal object of a partially ordered set is its maximal element, if one exists.
	\end{enumerate}
\end{example}

Again, terminal objects are unique up to isomorphism.
The reader is invited to repeat the proof from the preceding exercise here.
However, we can illustrate more strongly the notion of duality to give a short proof.
\begin{ques}
	Verify that terminal objects of $\AA$ are equivalent to initial objects of $\AA^{\text{op}}$.
	Thus terminal objects of $\AA$ are unique up to isomorphism.
\end{ques}
In general, one can consider in this way the dual of \emph{any} categorical notion:
properties of $\AA$ can all be translated to dual properties of $\AA^{\text{op}}$
(often by adding the word ``co'' in front).

One last neat construction: suppose we're working in a concrete category,
meaning (loosely) that the objects are ``sets with additional structure''.
Now suppose you're sick of maps and just want to think about elements of these sets.
Well, I won't let you do that since you're reading a category theory chapter,
but I will offer you the following advice:
\begin{itemize}
	\ii In $\catname{Set}$, arrows from $\{\ast\}$ to $S$ correspond to elements of $S$.
	\ii In $\catname{Top}$, arrows from $\{\ast\}$ to $X$ correspond to points of $X$.
	\ii In $\catname{Gp}$, arrows from $\ZZ$ to $G$ correspond to elements of $G$.
	\ii In $\catname{Ring}$, arrows from $\ZZ[x]$ to $R$ correspond to elements of $R$.
\end{itemize}
and so on.
So in most concrete categories, you can think of elements as functions from special sets to the set in question.



\section{Functors}
\prototype{Forgetful functors. Also $-^\vee$ is a contravariant functor.}
In mathematics, we'd often like to take an object from one category and get an object from another category.
For example, given a group we might like to obtain a set;
given a set, we might like to consider the vector space whose basis is that set,
and so on.
The nice way to do that is using the notion of a functor:
category theorists are always asking ``what are the maps?'',
and so we can now think about maps between categories.

\begin{definition}
	Let $\AA$ and $\BB$ be categories.
	A \vocab{functor} $F$ takes every object of $A$ to an object of $B$
	and every arrow $A_1 \taking{f} A_2$ to an arrow $F(A_1) \taking{F(f)} F(A_2)$.
	You can picture this as follows.
	\begin{diagram}
		& A_1 & & B_1 & = F(A_1) & \\
		\AA \ni & \dTo^f & \rDotted^F & \dTo_{F(f)} && \in \BB \\
		& A_2 & & B_2 & = F(A_2) &
	\end{diagram}
	(I'll try to use dotted arrows for functors, which cross different categories, for emphasis.)
	It needs to satisfy the following requirements:
	\begin{itemize}
		\ii Identity arrows get sent to identity arrows:
		for each identity arrow $\id_A$, we have $F(\id_A) = \id_{F(A)}$.
		\ii The functor respects composition:
		if $A_1 \taking f A_2 \taking g A_3$ are arrows in $\AA$,
		then $F(g \circ f) = F(g) \circ F(f)$.
	\end{itemize}
\end{definition}
\begin{example}
	[Free and Forgetful Functors]
	Note that these are both informal terms.
	\begin{enumerate}[(a)]
		\ii A \vocab{forgetful functor} is a functor that ``forgets'' some structure.
		For example, there is a forgetful functor $U : \catname{Vect}_k \to \catname{Set}$
		by taking $V$ to the set of vectors $U(V)$; certainly a linear map $T : V \to W$
		gives a map $U(T) : U(V) \to U(W)$ of the underlying point-sets.
		In other words we ``forget'' that $V$ has a vector space structure.

		Similarly there are forgetful functors from $\catname{Gp}$, $\catname{Ring}$, etc. to set.
		There is even a forgetful functor $\catname{Ring} \to \catname{Gp}$: send a ring $R$
		to the abelian group $(R,+)$.

		\ii A \vocab{free functor} is in some sense the opposite.
		A free functor $F : \catname{Set} \to \catname{Vect}_k$ can be taken by considering
		$F(S)$ to be the vector space with basis $S$.
	\end{enumerate}
\end{example}
\begin{remark}
	We can actually use this to give a definition of concrete category:
	it's a category with an ``injective'' forgetful functor to $U: \AA \to \catname{Set}$
	(The correct definition of ``injective'' here is \vocab{faithful}: no two arrows in $\AA$
	get sent to the same arrow in $\catname{Set}$.)
\end{remark}

\begin{example}
	[Functors from $\mathcal G$]
	Let $G$ be a group and $\mathcal G = \{\ast\}$ be the associated one-object category.
	\begin{enumerate}[(a)]
		\ii Consider a functor $F : \mathcal G \to \catname{Set}$, and let $S = F(\ast)$.
		Then the data of $F$ corresponds to putting a \emph{group action} of $G$ on $S$.
		\ii Consider a functor $F : \mathcal G \to \catname{FDVect}_k$, and let $V = F(\ast)$ have dimension $n$.
		Then the data of $F$ corresponds to embedding $G$ as a subgroup of the $n \times n$ matrices
		(i.e.\ the linear maps $V \to V$).
		This is one way groups historically arose; the theory of viewing groups as matrices
		forms the field of representation theory.
		\ii Let $H$ be a group and construct $\mathcal H$ the same way.
		Then functors $\mathcal G \to \mathcal H$ correspond to homomorphism $G \to H$.
	\end{enumerate}
\end{example}
\begin{exercise}
	Check the above group-based functors work as advertised.
\end{exercise}
%Now we present one much trickier example.
%If you don't get this one at first, skip it and come back later.
%\begin{example}
%	[Covariant Yoneda Functor]
%	For a category $\AA$, define the \vocab{covariant Yoneda functor} $H^A : \AA \to \catname{Set}$
%	by defining \[ H^A(A_1) \defeq \Hom_\AA (A, A_1) \in \catname{Set}. \]
%	Hence each $A_1$ is sent to the \emph{arrows from $A$ to $A_1$};
%	so \textbf{$H^A$ describes how $A$ sees the world}.
%	Now we want to specify how $H^A$ behaves on arrows.
%	For each arrow $A_1 \taking{f} A_2$, we need
%	to specify $\catname{Set}$-map $\Hom_\AA (A, A_1) \to \Hom(A, A_2)$;
%	in other words, we need to send an arrow $A \taking{p} A_1$ to an arrow $A \to A_2$.
%	There's only one reasonable way to do this: take the composition
%	\[ A \taking{p} A_1 \taking{f} A_2. \]
%	In other words, $H_A(f)$ is $p \mapsto p \circ f$.
%	In still other words, $H_A(f) = - \circ f$;
%	the $-$ is a slot for the input to go into.
%\end{example}

As another example:
\begin{ques}
	If $\mathbb P$ and $\mathbb Q$ are posets interpreted as categories,
	what does a functor from $\mathbb P$ to $\mathbb Q$ represent?
\end{ques}
\begin{exercise}
	Verify that there is a category of \emph{all categories},
	$\catname{Cat}$, whose arrows are functors.
	In particular, you will need to name the ``identity functors''
	and verify that the composition of a functor is also a functor.
\end{exercise}

The dual notion of a functor is a \vocab{contravariant functor}.
They work exactly the same way, except the arrows get flipped by the functor.
You can think of this is a functor $\AA \to \BB^{\text{op}}$ (equivalently, $\AA^{\text{op}} \to \BB$). Pictorially:
\begin{diagram}
	& A_1 & & B_1 & = F(A_1) & \\
	\AA \ni & \dTo^f & \rDotted^F & \uTo_{F(f)} && \in \BB \\
	& A_2 & & B_2 & = F(A_2) &
\end{diagram}
For emphasis, a usual functor is often called a \vocab{covariant functor}.

Contravariant functors come up in a lot in geometric applications.
Here's why.
If $X$ is a geometric object, we'll often consider
the \emph{set of functions} $X \taking\psi A$ for some particular $A$.
For example, if $V$ was a vector space, we could consider the functions $V \to k$,
giving the dual module $V^\vee$.
Or if $X$ was a space, we might consider the continuous
real functions $X \taking{p} \RR$.
As a non-geometric example: for a set $S$,
a function $S \to \{x,y\}$ corresponds to a subset of $S$.

Now suppose you have two spaces $X$, $Y$ and a map $f : X \to Y$.
Then given $Y \taking p \RR$, you automatically get a map $X \taking{f} Y \taking{p} \RR$.
And it turns out this is a functor!
This general construction can be done in any category $\AA$, as follows:
\begin{example}[Contravariant Yoneda Functor]
	The \vocab{contravariant Yoneda functor} $H_A : \AA^{\text{op}} \to \catname{Set}$;
	is used to describe how objects of $\AA$ see $A$.
	For each $X \in \AA$ it puts \[ H_A(X) \defeq \Hom_{\AA}(X, A) \in \catname{Set}. \]
	For $X \taking{f} Y$, $H_A(f)$ sends each arrow $Y \taking{p} A \in \Hom_\AA(Y,A)$ to 
	\[ X \taking{f} Y \taking{p} A \quad \in \Hom_\AA(X,A) \]
	as we did above.
	Thus $H_A(f)$ is an arrow from $\Hom_\AA(Y,A) \to \Hom_\AA(X,A)$.
	(Note the flipping!)
\end{example}
\begin{remark}
	Dually, there is also a \vocab{covariant Yoneda functor} $H^A : \AA \to \catname{Set}$
	by $H^A(X) = H^A(A, X)$. This guy describes how $A$ sees the world.
	\label{def:covariant_yoneda}
\end{remark}

\begin{example}[$V \mapsto V^\vee$ is Contravariant]
	Consider the category $\catname{Vect}_k$.
	If you read the digression on dual spaces,
	you now know there is a contravariant functor $\catname{Vect}_k \to \catname{Vect}_k$ by
	taking the \emph{dual} $V^\vee$ of a vector space $V$:
	\begin{diagram}
		V & & V^\vee \\
		\dTo^T & \rDotted^{-^\vee} & \uTo_{T^\vee} \\
		W & & W^\vee
	\end{diagram}
\end{example}

Suppose $G \cong H$ are groups, and $U : \catname{Gp} \to \catname{Set}$ is forgetful.
Then we get two sets $U(G)$ and $U(H)$.
These two sets have the same cardinality and are isomorphic in $\catname{Set}$.
Now the beauty of category theory shows itself: this in fact works \emph{for any functors and categories},
and the proof is done solely through arrows.

\begin{theorem}
	[Functors Preserve Isomorphism]
	\label{thm:functor_isom}
	If $A_1 \cong A_2$ are isomorphic objects in $\AA$
	and $F : \AA \to \BB$ is a functor
	then $F(A_1) \cong F(A_2)$.
\end{theorem}
\begin{proof}
	Try it yourself! The picture is:
	\begin{diagram}
		& A_1 & & B_1 & = F(A_1) & \\
		\AA \ni & \dTo^f \uTo_g & \rDotted^F & \dTo^{F(f)} \uTo_{F(g)} && \in \BB \\
		& A_2 & & B_2 & = F(A_2) &
	\end{diagram}
	You'll need to use both key properties of functors: they preserve composition
	and the identity map.
\end{proof}

This will be super convenient in algebraic topology, for example.
We'll show from a space $X$ you can create a group $\pi_n(X)$ or $H_n(X)$ that
gives information about the space $X$.
If we can show verify this is in fact a functor $\catname{Top} \to \catname{Gp}$,
then for $X \cong Y$, we get for free that $\pi_n(X) \cong \pi_n(Y)$ and $H_n(X) \cong H_n(Y)$.

\section{Natural Transformations}
We made categories to keep track of objects and maps, then went a little crazy and asked
``what are the maps between categories?'' to get functors.
Now we'll ask ``what are the maps between functions?'' to get natural transformations.

It might sound terrifying that we're drawing arrows between functors, but this is actually an old idea.
Recall that given two paths $\alpha, \beta : [0,1] \to X$,
we built a path-homotopy by ``continuously deforming'' the path $\alpha$ to $\beta$;
this could be viewed as a function $[0,1] \times [0,1] \to X$.
The definition of a natural transformation is similar: we want to pull $F$ to $G$
along a series of arrows in the target space $\BB$.

\begin{definition}
	Let $F, G : \AA \to \BB$ be two functors.
	A \vocab{natural transformation} $\alpha$ from $F$ to $G$, denoted
	\[ \nattfm{\AA}{F}{\alpha}{G}{\BB} \]
	consists of, for each $A \in \AA$ an arrow $\alpha_A \in \Hom_\BB(F(A), G(A))$, which is
	called the \vocab{component} of $\alpha$ at $A$.
	Pictorially, it looks like this:
	\begin{diagram}
		& & & F(A) \in \BB \\
		\AA \ni & A & \ruDotted(2,1)^F & \dTo_{\alpha_A} \\
		& & \rdDotted(2,1)^G & G(A) \in \BB
	\end{diagram}
	These $\alpha_A$ are subject to the ``naturality'' requirement that for any $A_1 \taking{f} A_2$,
	the diagram
	\begin{diagram}
		F(A_1) & \rTo^{F(f)} & F(A_2) \\
		\dTo_{\alpha_{A_1}} & & \dTo_{\alpha_{A_2}} \\
		G(A_1) & \rTo_{G(f)} & G(A_2)
	\end{diagram}
	commutes.
\end{definition}
The arrow $\alpha_A$ represents the path that $F(A)$ takes to get to $G(A)$
(just as in a path-homotopy from $\alpha$ to $\beta$
each \emph{point} $\alpha(t)$ gets deformed to the \emph{point} $\beta(t)$ continuously).
A picture might help: consider
\begin{center}
	\begin{asy}
		size(14cm);
		dotfactor *= 1.4;

		path sparrow(pair X, pair Y) {
			// Short for "spaced arrow"
			return (0.9*X+0.1*Y)--(0.1*X+0.9*Y);
		}

		pair A1 = Drawing("A_1", dir(210), dir(225));
		pair A2 = Drawing("A_2", origin, dir(90));
		pair A3 = Drawing("A_3", dir(-30), dir(-45));
		path f = Drawing(sparrow(A2, A1), EndArrow);
		label("$f$", f, dir(90));
		path g = Drawing(sparrow(A2, A3), EndArrow);
		label("$g$", g, dir(90));
		label("$\mathcal A$", 0.6*(A1+A3));

		pen p = blue;
		transform FF = shift( (3.5, 0.7) );
		dot("$F(A_1)$", FF*A1, dir(225), p);
		dot("$F(A_2)$", FF*A2, dir(90), p);
		dot("$F(A_3)$", FF*A3, dir(-45), p);
		draw(FF*f, p, EndArrow);
		draw(FF*g, p, EndArrow);
		label("$F(f)$", FF*f, dir(110), p);
		label("$F(g)$", FF*g, dir(70), p);
		draw(FF*f, p+1.4);
		draw(FF*g, p+1.4);

		p = deepcyan;
		transform GG = shift( (3.5, -0.7) );
		dot("$G(A_1)$", GG*A1, dir(225), p);
		dot("$G(A_2)$", GG*A2, 3*dir(-90), p);
		dot("$G(A_3)$", GG*A3, dir(-45), p);
		label("$G(f)$", D(GG*f, p, EndArrow), dir(110), p);
		label("$G(g)$", D(GG*g, p, EndArrow), dir(70), p);
		draw(GG*f, p+1.4);
		draw(GG*g, p+1.4);

		p = lightred;
		label("$\alpha_{A_1}$", D(sparrow(FF*A1, GG*A1), p, EndArrow), dir(180), p);
		label("$\alpha_{A_2}$", D(sparrow(FF*A2, GG*A2), p, EndArrow), dir(180), p);
		label("$\alpha_{A_3}$", D(sparrow(FF*A3, GG*A3), p, EndArrow), dir(0), p);

		p = magenta + dotted + 0.7;
		path Fa = (0.5,0)--FF*(-1,-0.2);
		path Ga = (0.5,-0.6)--GG*(-1,-0.4);
		label("$F$", D(Fa, p, EndArrow), dir(135), p);
		label("$G$", D(Ga, p, EndArrow), dir(225), p);

		p = lightred + 0.7;
		label("$\alpha$", D(sparrow(midpoint(Fa), midpoint(Ga)), p, EndArrow), dir(180), p);

		p = grey + dashed;
		pointpen = p;
		pair B1 = D(midpoint(FF*A2--GG*A1));
		pair B2 = D(0.6 * (FF*A3) + 0.4 * (GG*A2));
		draw(sparrow(FF*A1, B1), p, EndArrow);
		draw(sparrow(GG*A2, B1), p, EndArrow);
		draw(sparrow(FF*A3, B2), p, EndArrow);
		pair B3 = D(FF*A3 + 0.7*dir(100));
		draw(sparrow(B3, FF*A3), p, EndArrow);
		label("$\mathcal B$", GG*(0.6*(A1+A3)));
		draw(sparrow(FF*A2, GG*A3), p, EndArrow);
		pair B4 = D(FF*A1 + 0.5*dir(90));
		draw(sparrow(FF*A1, B4), p, EndArrow);
	\end{asy}
\end{center}
Here $\AA$ is the small black category with three elements and two non-identity arrows $f$, $g$.
(I've omitted the identity arrows for simplicity).
The images of $\AA$ under $F$ and $G$ are the blue and green ``subcategories'' of $\BB$.
Note that $\BB$ could potentially have many more objects and arrows in it (grey).
The natural transformation $\alpha$ (red) selects an arrow of $\BB$ from each $F(A)$
to the corresponding $G(A)$, such dragging the entire image of $F$ to the image of $G$.
Finally, we require that any diagram formed by the blue, red, and green arrows is commutative (naturality),
so the natural transformation is really ``natural''.

There is a second equivalent definition that looks much more like the homotopy.
\begin{definition}
	Let $\mathbf 2$ denote the category generated by a poset with two elements $0 \le 1$, that is,
	\begin{center}
	\begin{tikzpicture}[scale=2]
		\SetVertexMath
		\Vertices{circle}{1,0}
		\Edge[style={->}, label={$0 \le 1$}](0)(1)
		\Loop[dist=12, dir=NO, label={$\id_0$}, labelstyle={above=1pt}](0)
		\Loop[dist=12, dir=NO, label={$\id_1$}, labelstyle={above=1pt}](1)
	\end{tikzpicture}
	\end{center}
	Then a \emph{natural transformation} 
	$ \nattfm{\AA}{F}{\alpha}{G}{\BB} $
	is just a functor $\alpha : \AA \times \mathbf 2 \to \BB$ with
	\[ \alpha(0,A) = F(A), \;\; \alpha(0,f) = F(f) 
		\quad\text{and}\quad
	\alpha(1,A) = G(A), \;\; \alpha(1,f) = G(f). \]
	or more succinctly, $\alpha(0,-) = F$, $\alpha(1,-) = G$.
\end{definition}
The proof that these are equivalent is left as a practice problem.

Naturally, two natural transformations $\alpha : F \to G$ and $\beta : G \to H$ can get composed.
\begin{diagram}
	& & F(A) \\
	& \ruDotted^F & \dTo_{\alpha_A} \\
	\AA \ni A & \rDotted^G & G(A) \\
	& \rdDotted_H & \dTo_{\beta_A} \\
	&& H(A)
\end{diagram}

Now suppose $\alpha$ is a natural transformation such that $\alpha_A$ is an isomorphism for each $A$.
In this way, we can construct an inverse arrow $\beta_A$ to it.
\begin{diagram}
	& & F(A) \in \BB \\
	\AA \ni A & \ruDotted(2,1)^F & \dTo^{\alpha_A} \uTo_{\beta_A} \\
	& \rdDotted(2,1)^G & G(A) \in \BB
\end{diagram}
In this case, we say $\alpha$ is a \vocab{natural isomorphism}.
We can then say that $F(A) \cong G(A)$ \vocab{naturally} in $A$.
(And $\beta$ is an isomorphism too!)

This is what it really means when we say that ``there is a natural / canonical isomorphism''.
For example, I claimed earlier that there was a canonical isomorphism $(V^\vee)^\vee \cong V$,
and mumbled something about ``not having to pick a basis'' and ``God-given''.
Category theory, amazingly, lets us formalize this:
it just says that $(V^\vee)^\vee \cong \id(V)$ naturally in $V \in \catname{FDVect}_k$.
Really, we have a natural transformation
\[ \nattfm{\catname{FDVect}_k}{\id}{\eps}{(-^\vee)^\vee}{\catname{FDVect}_k}. \]
where the component $\eps_V$ is given by $v \mapsto \opname{ev}_v$
(as discussed earlier,
the fact that it is an isomorphism follows from the fact that $V$ and $(V^\vee)^\vee$
have equal dimensions and $\eps_V$ is injective).

% A last remark: similar to the situation with groups isomorphisms,
% collecting all these $\beta_A$'s gives a natural isomorphism too.
\begin{ques}
	In fact, there is a category $[\AA, \BB]$ of \emph{functors} from $\AA$ into $\BB$,
	whose arrows are natural transformations.
	When are two functors isomorphic in $[\AA, \BB]$?
\end{ques}
As a result of this, we also write $F \cong G$ to mean $F(X) \cong G(X)$

\section\problemhead
\begin{dproblem}
	Let $X$ and $Y$ be objects in a category $\AA$.
	Show that $X \cong Y$ if and only if $H_X(A) \cong H_Y(A)$ naturally in $A$.\footnote{
		From \cite{ref:msci}: let's consider $\AA = \catname{Gp}$ for concreteness.
		Suppose $A$, $X$, $Y$ are groups such that $H_X(A) \cong H_Y(A)$ for all $A$.
		If $A = \ZZ$, then $\left\lvert X \right\rvert = \left\lvert Y \right\rvert$.
		If $A = \ZZ / p\ZZ$, then $X$ and $Y$ have the same number of elements of order $p$.
		Each $A$ gives us some information on how $X$ and $Y$ are similar,
		but the whole natural isomorphism is strong enough to imply $X \cong Y$.}
\end{dproblem}


\begin{problem}
	Show that the two definitions of natural transformation
	(one in terms of $\AA \times \mathbf 2 \to \BB$
	and one in terms of arrows $F(A) \taking{\alpha_A} G(A)$) 
	are equivalent.
	\begin{hint}
		The category $\AA \times \mathbf 2$ has ``redundant arrows''.
	\end{hint}
	\begin{sol}
		The main observation is that in $\AA \times \mathbf 2$,
		you have the arrows in $\AA$ (of the form $(f, \id_{\mathbf 2})$),
		and then the arrows crossing the two copies of $\AA$ (of the form $(\id_A, 0 \le 1)$).
		But there are some more arrows $(f, 0 \le 1)$: nonetheless, they can be thought of as compositions
		\[ (f, 0 \le 1) = (f, \id_{\mathbf 2}) \circ (\id_A, 0 \le 1) = (\id_A, 0 \le 1) \circ (f, \id_{\mathbf 2}). \]
		Now we want to specify a functor $\alpha : \AA \times \mathbf 2$, we only have to specify
		where each of these two more basic things goes.
		The conditions on $\alpha$ already tells us that $(f, \id_{\mathbf 2})$ should be mapped to $F(f)$ or $G(f)$
		(depending on whether the arrow above is in $\AA \times \{0\}$ or $\AA \times \{1\}$),
		and specifying the arrow $(\id_A, 0 \le 1)$ amounts to specifying the $A$th component.
		Where does naturality come in?

		The above discussion transfers to products of categories in general:
		you really only have to think about $(f, \id)$ and $(\id, g)$ arrows
		to get the general arrow $(f,g) = (f, \id) \circ (\id, g) = (\id, g) \circ (f, \id)$.
	\end{sol}
\end{problem}

\begin{problem}
	Let $\AA$ be the category of finite sets whose arrows are bijections between sets.
	For $A \in \AA$,
		let $F(A)$ be the set of \emph{permutations} of $A$ and
		let $G(A)$ be the set of \emph{orderings} on $A$.\footnote{
			A permutation is a bijection $A \to A$,
			and an ordering is a bijection $\{1, \dots, n\} \to A$,
			where $n$ is the size of $A$.}
	\begin{enumerate}[(a)]
		\ii Extend $F$ and $G$ to functors $\AA \to \catname{Set}$.
		\ii Show that $F(A) \cong G(A)$ for every $A$, but this isomorphism is \emph{not} natural.
	\end{enumerate}
\end{problem}

\begin{problem}
	[Yoneda Lemma]
	Let $\AA$ be a category, pick $A \in \AA$, and let $H_A$ be the contravariant Yoneda functor.
	\begin{enumerate}[(a)]
	\ii \gim
	Let $X : \AA^{\text{op}} \to \catname{Set}$ be a contravariant functor.
	Exhibit a bijection
	\[ X(A) \cong \left\{ \text{Natural transformations }
		\nattfm{\AA^{\text{op}}}{H_A}{\alpha}{X}{\catname{Set}}
	\right\}. \]
	\ii \kurumi Moreover, interpreting both sides as functors
	$[\AA^{\text{op}}, \catname{Set}] \times \AA^{\text{op}} \to \catname{Set}$
	show that the isomorphism is \emph{natural} in $X$ and $A$.
	\end{enumerate}
\end{problem}
The Yoneda Lemma has tons of very cool consequences; one is that solving it
makes you feel like a minor god.
The proof is long, but it is not actually very difficult -- from start to finish,
there is only one thing you can possibly do.
