\chapter{Algebraic integers}
Here's a first taste of algebraic number theory.

This is really close to the border between olympiads and higher math.
You've always known that $a+\sqrt2 b$ had a ``norm'' $a^2-2b^2$,
and that somehow this norm was multiplicative.
You've also always known that roots come in conjugate pairs.
You might have heard of minimal polynomials but not know much about them.

This chapter will make all these vague notions precise.
It's drawn largely from the first chapter of \cite{ref:oggier_NT}.

\section{Motivation from high school algebra}
This is adapted from my blog, \emph{Power Overwhelming}\footnote{URL: \url{https://usamo.wordpress.com/2014/10/19/why-do-roots-come-in-conjugate-pairs/}}.

In high school precalculus, you'll often be asked to find the roots of some polynomial with integer coefficients.
For instance,
\[ x^3 - x^2 - x - 15 = (x-3)(x^2+2x+5) \]
has roots $3$, $-1+2i$, $-1-2i$.
Or as another example, 
\[ x^3 - 3x^2 - 2x + 2 = (x+1)(x^2-4x+2) \]
has roots $-1$, $2 + \sqrt 2$, $2 - \sqrt 2$.
You'll notice that the irrational roots, like $-1 \pm 2i$ and $2 \pm \sqrt 2$, are coming up in pairs. In fact, I think precalculus explicitly tells you that the imaginary roots come in conjugate pairs. More generally, it seems like all the roots of the form $a + b \sqrt c$ come in ``conjugate pairs''. And you can see why.

But a polynomial like
\[ x^3 - 8x + 4 \]
has no rational roots.
(The roots of this are approximately $-3.0514$, $0.51730$, $2.5341$.)
Or even simpler,
\[ x^3 - 2 \]
has only one real root, $\sqrt[3]{2}$.
These roots, even though they are irrational, have no ``conjugate'' pairs.
Or do they?

Let's try and figure out exactly what's happening.
Let $\alpha$ be any complex number.
We define the \vocab{minimal polynomial} of $\alpha$ over $\QQ$ to be the polynomial
\begin{itemize}
	\item $P(x)$ has rational coefficients, and leading coefficient $1$,
	\item $P(\alpha) = 0$.
	\item The degree of $P$ is as small as possible.
		This is called the \vocab{degree} of $\alpha$.
\end{itemize}
\begin{example}[Examples of minimal polynomials]
	\listhack
	\begin{enumerate}[(a)]
	\ii $\sqrt 2$ has minimal polynomial $x^2-2$.
	\ii The imaginary unit $i = \sqrt{-1}$ has minimal polynomial $x^2+1$.
	\ii A primitive $p$th root of unity, $\zeta_p = e^{\frac{2\pi i}{p}}$, has minimal polynomial $x^{p-1} + x^{p-2} + \dots + 1$.
	\end{enumerate}
\end{example}
Note that $100x^2 - 200$ is also a polynomial of the same degree which has $\sqrt 2$ as a root; that's why we want to require the polynomial to be monic. That's also why we choose to work in the rational numbers; that way, we can divide by leading coefficients without worrying if we get non-integers.

Why do we care? The point is as follows: suppose we have another polynomial $A(x)$ such that $A(\alpha) = 0$.
Then we claim that $P(x)$ actually divides $A(x)$!
That means that all the other roots of $P$ will also be roots of $A$.

The proof is by contradiction: if not, by polynomial long division we can find a quotient and remainder $Q(x)$, $R(x)$ such that
\[ A(x) = Q(x) P(x) + R(x) \]
and $R(x) \not\equiv 0$.
Notice that by plugging in $x = \alpha$, we find that $R(\alpha) = 0$.
But $\deg R < \deg P$, and $P(x)$ was supposed to be the minimal polynomial.
That's impossible!

Let's look at a more concrete example.
Consider $A(x) = x^3-3x^2-2x+2$ from the beginning. 
The minimal polynomial of $2 + \sqrt 2$ is $P(x) = x^2 - 4x + 2$ (why?).
Now we know that if $2 + \sqrt 2$ is a root, then $A(x)$ is divisible by $P(x)$.
And that's how we know that if $2 + \sqrt 2$ is a root of $A$, then $2 - \sqrt 2$ must be a root too.

As another example, the minimal polynomial of $\sqrt[3]{2}$ is $x^3-2$. So $\sqrt[3]{2}$ actually has \textbf{two} conjugates, namely, $\alpha = \sqrt[3]{2} \left( \cos 120^\circ + i \sin 120^\circ \right)$ and $\beta = \sqrt[3]{2} \left( \cos 240^\circ + i \sin 240^\circ \right)$. Thus any polynomial which vanishes at $\sqrt[3]{2}$ also has $\alpha$ and $\beta$ as roots!

\begin{ques}
	Let $\alpha$ be a root of the polynomial $P(x)$.
	Show that $P(x)$ is the minimal polynomial if and only if it is irreducible.
	(This is tautological: the point is just to realize that ``minimal polynomials'' and ``irreducible polynomials'' are the same beasts.)
\end{ques}

\section{Algebraic numbers and algebraic integers}
\prototype{$\sqrt2$ is an algebraic integer (root of $x^2-2$),
$\half$ is an algebraic number but not an algebraic integer (root of $x-\half$).}
Let's now work in much vaster generality.
First, let's give names to the new numbers we've discussed above.
\begin{definition}
	An \vocab{algebraic number} is any $\alpha \in \CC$
	which is the root of \emph{some} polynomial with coefficients in $\QQ$.
	The set of algebraic numbers is denoted $\ol\QQ$.
\end{definition}
\begin{remark}
	One can equally well say algebraic numbers are those of which 
	are roots of some polynomial with coefficients in $\ZZ$ (rather than $\QQ$),
	since any polynomial in $\QQ[x]$ can be scaled to one in $\ZZ[x]$.
\end{remark}
\begin{definition}
	Consider an algebraic number $\alpha$ and its minimal polynomial $P$ (which is monic and has rational coefficients).
	If it turns out the coefficients of $P$ are integers, then we say $\alpha$ is an \vocab{algebraic integer}.

	The set of algebraic integers is denoted $\ol\ZZ$.
\end{definition}
\begin{remark}
	One can show, using \emph{Gauss's Lemma}, that if $\alpha$ is the root
	of \emph{any} monic polynomial with integer coefficients, then $\alpha$ is an algebraic integer.
	So in practice, if I want to prove that $\sqrt 2 + \sqrt 3$ is an algebraic integer,
	then I only have to say ``the polynomial $(x^2-5)^2-6$ works'' without checking that it's minimal.
\end{remark}
Sometimes for clarity, we refer to elements of $\ZZ$
as \vocab{rational integers}.
\begin{example}[Examples of algebraic integers]
	The numbers
	\[ 4, \; i = \sqrt{-1}, \; \sqrt[3]{2}, \; \sqrt2+\sqrt3 \]
	are all algebraic integers, since they are the roots of the monic polynomials
	$x-4$, $x^2+1$, $x^3-2$ and $(x^2-5)^2-6$.

	The number $\half$ has minimal polynomial $x - \half$,
	so it's an algebraic number but not an algebraic integer.
	(In fact, the rational root theorem also directly implies
	that any monic integer polynomial does not have $\half$ as a root!)
\end{example}
\begin{proposition}[Rational algebraic integers are rational integers]
	An algebraic integer is rational if and only if it is a rational integer.
	In symbols, \[ \ol\ZZ \cap \QQ = \ZZ. \]
\end{proposition}
\begin{ques}
	(Important) Prove this.
\end{ques}

\section{Number fields}
\prototype{$\QQ(\sqrt2)$ is a typical number field.}
Given any algebraic number $\alpha$, we're able to consider fields of the form $\QQ(\alpha)$. Let us write down the more full version.
\begin{definition}
	A \vocab{number field} $K$ is a field containing $\QQ$ as a subfield
	which is a \emph{finite-dimensional} $\QQ$-vector space.
	The \vocab{degree} of $K$ is its dimension.
\end{definition}
\begin{example}[Prototypical example]
	Consider the field
	\[ K = \QQ(\sqrt2) = \left\{ a+b\sqrt2 \mid a,b \in \QQ \right\}. \]
	This is a field extension of $\QQ$,
	and has degree $2$ (the basis being $1$ and $\sqrt2$).
\end{example}
\begin{example}[Adjoining an algebraic number]
	Let $\alpha$ be the root of some irreducible polynomial $P(x)$ in $\QQ$.
	The field $\QQ(\alpha)$ is a field extension as well, and the basis
	is $1, \alpha, \alpha^2, \dots, \alpha^{m-1}$,
	where $m$ is the degree of $\alpha$.
	In particular, the degree of $\QQ(\alpha)$ is just the degree of $\alpha$.
\end{example}
\begin{example}[Non-examples of number fields]
	$\RR$ and $\CC$ are not number fields since there is no \emph{finite}
	$\QQ$-basis of them.
\end{example}

\section{Norms and traces}
\prototype{$a+b\sqrt2$ as an element of $\QQ(\sqrt2)$ has norm $a^2-2b^2$ and trace $2a$.}
Remember when you did olympiads and we had like $a^2+b^2$ was the ``norm'' of $a+bi$?
Cool, let me tell you what's actually happening.

First, let me make precise the notion of a conjugate.
\begin{definition}
	Let $\alpha$ be an algebraic number, and let $P(x)$ be its minimal polynomial,
	of degree $m$.
	Then the $m$ roots of $P$ are the (Galois) \vocab{conjugates} of $\alpha$.
\end{definition}
It's worth showing at the moment that there are no repeated conjugates.
\begin{lemma}[Irreducible Polynomials Have Distinct Roots]
	An irreducible polynomial cannot have a complex double root.
	\label{lem:irred_complex}
\end{lemma}
\begin{proof}
	Let $f(x) \in \QQ[x]$ be the irreducible polynomial and assume it has a double root $\alpha$.
	\textbf{Take the derivative $f'(x)$.}
	This derivative has three interesting properties.
	\begin{itemize}
		\ii The degree of $f'$ is one less than the degree of $f$.
		\ii The polynomials $f$ and $f'$ are not relatively prime
		because they share a factor $x-\alpha$.
		\ii The coefficients of $f'$ are also in $\QQ$.
	\end{itemize}
	Consider $P = \gcd(f, f')$. We must have $P \in \QQ[x]$ by Euclidean algorithm.
	But the first two facts about $f'$ ensure that $P$ is nonconstant
	and $\deg P < \deg f$. Yet $P$ divides $f$, contradiction.
\end{proof}
Hence $\alpha$ has exactly as many conjugates as the degree of $\alpha$.

Now, we would \emph{like} to define the \emph{norm} of an element $\Norm(\alpha)$
as the product of its conjugates.
For example, we want $2+i$ to have norm $(2+i)(2-i) = 5$,
and in general for $a+bi$ to have norm $a^2+b^2$.
It would be \emph{really cool} if the norm was multiplicative;
we already know this is true for complex numbers!

Unfortunately, this doesn't quite work: consider
\[ \Norm(2+i) = 5 \text{ and } \Norm(2-i) = 5. \]
But $(2+i)(2-i) = 5$, which doesn't have norm $25$ like we want,
since $5$ is degree $1$ and has no conjugates at all.
The reason this ``bad'' thing is happening is that we're
trying to define the norm of an \emph{element},
when we really ought to be defining the norm of an element
\emph{with respect to a particular $K$}.

What I'm driving at is that the norm should have different meanings depending on which field you're in.
If we think of $5$ as an element of $\QQ$, then its norm is $5$.
But thought of as an element of $\QQ(i)$, its norm really ought to be $25$.
Let's make this happen: for $K$ a number field, we will now define $\Norm_{K/\QQ}(\alpha)$
to be the norm of $\alpha$ \emph{with respect to $K$} as follows.
\begin{definition}
	Let $\alpha \in K$ have degree $n$, so $\QQ(\alpha) \subseteq K$, and set $k = (\deg K) / n$.
	The \vocab{norm} of $\alpha$ is defined as
	\[ \Norm_{K/\QQ}(\alpha) \defeq \left( \prod \text{Galois conj of $\alpha$} \right)^k. \]
	The \vocab{trace} is defined as
	\[ \Tr_{K/\QQ}(\alpha) \defeq k \cdot \left( \sum \text{Galois conj of $\alpha$} \right). \]
\end{definition}
The exponent of $k$ is a ``correction factor'' that makes the norm of $5$ into $5^2=25$
when we view $5$ as an element of $\QQ(i)$ rather than an element of $\QQ$.
For a ``generic'' element of $K$, we expect $k = 1$.
\begin{exercise}
	Use what you know about nested vector spaces to convince
	yourself that $k$ is actually an integer.
\end{exercise}
\begin{example}[Norm of $a+b\sqrt2$]
	Let $\alpha = a+b\sqrt2 \in \QQ(\sqrt2)$.
	If $b \neq 0$, then $\alpha$ and $K$ have the degree $2$.
	Thus the only conjugate of $\alpha$ is $a-b\sqrt2$, which gives
	the norm \[ (a+b\sqrt2)(a-b\sqrt2) = a^2-2b^2, \]
	The trace is $(a-b\sqrt2) + (a+b\sqrt2) = 2a$.

	Nicely, the formula $a^2-2b^2$ and $2a$ also works when $b=0$.
\end{example}
Of importance is the following.
\begin{proposition}[Norms and traces are rational integers]
	If $\alpha$ is an algebraic integer, its norm and trace are traces 
	are rational integers.
\end{proposition}
\begin{ques}
	Prove it. (Vieta.)
\end{ques}

That's great, but it leaves a question unanswered:
why is the norm multiplicatve?
To do this, I have to give a new definition of norm and trace.

\begin{theorem}[Morally correct definition of norm and trace]
	Let $K$ be a number field of degree $n$, and let $\alpha \in K$.
	Let $\mu_\alpha : K \to K$ denote the map \[ x \mapsto \alpha x \]
	viewed as a linear map of $\QQ$-vector spaces.
	Then,
	\begin{itemize}
		\ii the norm of $\alpha$ equals the determinant $\det \mu_\alpha$, and
		\ii the trace of $\alpha$ equals the trace $\Tr \mu_\alpha$.
	\end{itemize} 
\end{theorem}
Since the trace and determinant don't depend on the choice of basis,
you can pick whatever basis you want
and use whatever definition you got in high school.
Fantastic, right?

\begin{example}[Explicit computation of matrices for $a+b\sqrt2$]
	Let $K = \QQ(\sqrt2)$, and let $1$, $\sqrt 2$ be the basis of $K$.
	Let \[ \alpha = a + b \sqrt 2 \] (possibly even $b = 0$), and notice that
	\[ \left( a+b\sqrt2 \right)\left( x+y\sqrt2 \right)
		= (ax+2yb) + (bx+ay)\sqrt2. \]
	We can rewrite this in matrix form as
	\[ 
		\left(
		\begin{array}{cc}
			a & 2b \\
			b & a
		\end{array}
		\right)
		\left(
		\begin{array}{c}
			x \\
			y
		\end{array}
		\right)
		=
		\left(
		\begin{array}{c}
			ax+2yb \\
			bx+ay
		\end{array}
		\right).
	\]
	Consequently, we can interpret $\mu_\alpha$ as the matrix
	\[ \mu_\alpha = \left(
		\begin{array}{cc}
			a & 2b \\
			b & a
		\end{array}
		\right). \]
	Of course, the matrix will change if we pick a different basis,
	but the determinant and trace do not: they are always given by
	\[ \det \mu_\alpha = a^2-2b^2 \text{ and }
		\Tr \mu_\alpha = 2a. \]
\end{example}
This interpretation explains why the same formula should work for $a+b\sqrt 2$
even in the case $b = 0$.

\begin{proof}
I'll prove the result for just the norm; the trace falls out similarly.
Set
\[ n = \deg \alpha \quad kn = \deg K. \]
The proof is split into two parts, depending on whether or not $k=1$.
\begin{subproof}[Proof if $k=1$]
	Set $n = \deg \alpha = \deg K$.
	Thus the norm actually \emph{is} the product of the Galois conjugates.
	Also, \[ \{1, \alpha, \dots, \alpha^{n-1}\} \]
	is linearly independent in $K$, and hence a basis (as $\dim K = n$).
	Let's use this as the basis for $\mu_\alpha$.

	Let \[ x^n+c_{n-1}x^{n-1} + \dots + c_0  \]be the minimal polynomial of $\alpha$.
	Thus $\mu_\alpha(1) = \alpha$, $\mu_\alpha(\alpha) = \alpha^2$, and so on,
	but $\mu_\alpha(\alpha^{n-1}) = -c_{n-1}\alpha^{n-1} - \dots - c_0$.
	Therefore, $\mu_\alpha$ is given by the matrix
	\[
		M = 
		\left(
		\begin{array}{cccccc}
			0 & 0 & 0 & \dots & 0 & -c_0 \\
			1 & 0 & 0 & \dots & 0 & -c_1 \\
			0 & 1 & 0 & \dots & 0 & -c_2 \\
			\vdots & \vdots & \vdots & \ddots & 0 & -c_{n-2} \\
			0 & 0 & 0 & \dots & 1 & -c_{n-1}
		\end{array}
		\right).
	\]
	Thus \[ \det M = (-1)^n c_0 \] and we're done by Vieta's formulas.
\end{subproof}
\begin{subproof}[Proof if $k > 1$]
	We have nested vector spaces
	\[ \QQ \subseteq \QQ(\alpha) \subseteq K. \]
	Let $e_1$, \dots, $e_k$ be a $\QQ(\alpha)$-basis for $K$
	(meaning: interpret $K$ as a vector space over $\QQ(\alpha)$, and pick that basis).
	Since $\{1, \alpha, \dots, \alpha^{n-1}\}$ is a $\QQ$ basis for $\QQ(\alpha)$,
	the elements
	\[
		\begin{array}{cccc}
		e_1, & e_1\alpha, & \dots, & e_1\alpha^{n-1} \\
		e_2, & e_2\alpha, & \dots, & e_2\alpha^{n-1} \\
		\vdots & \vdots & \ddots & \vdots \\
		e_k, & e_k\alpha, & \dots, & e_k\alpha^{n-1}
		\end{array}
	\]
	constitute a $\QQ$-basis of $K$.
	Using \emph{this} basis, the map $\mu_\alpha$ looks like
	\[
		\left(
		\underbrace{
		\begin{array}{cccc}
			M & & & \\
			& M & & \\
			& & \ddots & \\
			& & & M
		\end{array}
		}_{\text{$k$ times}}
		\right)
	\]
	where $M$ is the same matrix as above:
	we just end up with one copy of our old matrix for each $e_k$.
	Thus $\det \mu_\alpha = (\det M)^k$, as needed. \qedhere
\end{subproof}
\begin{ques}
	Verify the result for traces as well. \qedhere
\end{ques}
\end{proof}

From this it follows immediately that 
\[ \Norm_{K/\QQ}(\alpha\beta) = \Norm_{K/\QQ}(\alpha)\Norm_{K/\QQ}(\beta) \]
because by definition we have 
\[ \mu_{\alpha\beta} = \mu_\alpha \circ \mu_\beta, \]
and that the determinant is multiplicative.
In the same way, the trace is additive.

\section{The ring of integers}
\prototype{If $K = \QQ(\sqrt 2)$, then $\OO_K = \ZZ[\sqrt 2]$.
	But if $K = \QQ(\sqrt 5)$, then $\OO_K = \ZZ[\frac{1+\sqrt5}{2}]$.}

$\ZZ$ makes for better number theory than $\QQ$.
In the same way, focusing on the \emph{algebraic integers} of $K$
gives us some really nice structure, and we'll do that here.

\begin{definition}
	Given a number field $K$, we define 
	\[ \OO_K \defeq K \cap \ol\ZZ \]
	to be the \vocab{ring of integers} of $K$;
	in other words $\OO_K$ consists of the algebraic integers of $K$.
\end{definition}

We do the classical example of a quadratic field now.
Before proceeding, I need to write a silly number theory fact.
\begin{exercise}
	Let $a$ and $b$ be rational numbers, and $d$ a squarefree positive integer.
	\begin{itemize}
		\ii If $d \equiv 2, 3 \pmod 4$, prove that
		$2a, a^2-db^2 \in \ZZ$ if and only if $a,b \in \ZZ$.
		\ii For $d \equiv 1 \pmod 4$, prove that
		$2a, a^2-db^2 \in \ZZ$ if and only if $a,b \in \ZZ$
		OR if $a -\half, b-\half \in \ZZ$.
	\end{itemize}
	This is annoying but not hard; you'll need to take mod $4$.
\end{exercise}

\begin{example}
	[Ring of integers of $K = \QQ(\sqrt3)$]
	Let $K$ be as above.
	We claim that \[ \OO_K = \ZZ[\sqrt 3] = \left\{ m + n\sqrt 3 \mid m,n \in \ZZ  \right\}. \]
	We set $\alpha = a + b \sqrt 3$.
	Then $\alpha \in \OO_K$ when the minimal polynomial has integer coefficients.

	If $b = 0$, then the minimal polynomial is $x-\alpha=x-a$,
	and thus $\alpha$ works if and only if it's an integer.
	If $b \neq 0$, then the minimal polynomial is
	\[ (x-a)^2 - 3b^2 = x^2 - 2a \cdot x + (a^2-3b^2). \]
	From the exercise, this occurs exactly for $a,b \in \ZZ$.
\end{example}
\begin{example}
	[Ring of integers of $K = \QQ(\sqrt 5)$]
	We claim that in this case
	\[ \OO_K = \ZZ\left[ \frac{1+\sqrt5}{2} \right]
	= \left\{ m + n \cdot \frac{1+\sqrt5}{2} \mid m,n \in \ZZ \right\}. \]
	The proof is exactly the same, except the exercise tells us instead
	that for $b \neq 0$, we have both the possibility that $a,b \in \ZZ$
	or that $a,b \in \ZZ - \half$.
	This reflects the fact that $\frac{1+\sqrt5}{2}$ is the root of $x^2-x-1 = 0$;
	no such thing is possible with $\sqrt 3$.
\end{example}
In general, the ring of integers of $K = \QQ(\sqrt d)$ is
\[ \OO_K 
	=
	\begin{cases}
		\ZZ[\sqrt d] & d\equiv 2,3 \pmod 4 \\[1em]
		\ZZ\left[ \frac{1+\sqrt d}{2} \right] & d \equiv 1 \pmod 4.
	\end{cases}
\]
What we're going to show is that $\OO_K$ behaves in $K$
a lot like the integers do in $\QQ$.
First we show $K$ consists of quotients of numbers in $\OO_K$.
In fact, we can do better:
\begin{example}[Rationalizing the denominator]
	For example, consider $K = \QQ(\sqrt3)$.
	The number $x = \frac{1}{4+\sqrt3}$ is an element of $K$, but by
	``rationalizing the denominator'' we can write
	\[ \frac{1}{4+\sqrt3} = \frac{4-\sqrt3}{13}. \]
	So we see that in fact, $x$ is $\frac{1}{13}$ of an integer in $\OO_K$.
\end{example}

The theorem holds true more generally.
\begin{theorem}[$K = \QQ \cdot \OO_K$]
	Let $K$ be a number field, and let $x \in K$ be any element.
	Then there exists an integer $n$ such that $nx \in \OO_K$;
	in other words, \[ x = \frac 1n \alpha \] for some $\alpha \in \OO_K$.
\end{theorem}
\begin{exercise}
	Prove this yourself.
	(Start by using the fact that $x$ has a minimal
	polynomial with integer coefficients.
	Alternatively, take the norm.)
\end{exercise}

Now we are show $\OO_K$ is a ring;
we'll check it is closed under addition and multiplication.
To do so, the easiest route is the following lemma.

\newcommand{\tempaeuqnxhj}{$\ZZ[\alpha]$} % fml latex
\begin{lemma}[$\alpha \in \ol\ZZ$ $\iff$ \tempaeuqnxhj\ finitely generated]
	Let $\alpha \in \ol\QQ$.
	Then $\alpha$ is an algebraic integer if and only if
	the abelian group $\ZZ[\alpha]$ is finitely generated.
\end{lemma}
\begin{proof}
	Note that $\alpha$ is an algebraic integer if and only if it's the root
	of some nonzero, monic polynomial with integer coefficients.
	Suppose first that
	\[ \alpha^N = c_{N-1} \alpha^{N-1} + c_{N-2} \alpha^{N-2} + \dots + c_0. \]
	Then the set $1, \alpha, \dots, \alpha^{N-1}$ generates $\ZZ[\alpha]$,
	since we can repeatedly replace $\alpha^N$ until all powers of $\alpha$
	are less than $N$.

	Conversely, suppose that $\ZZ[\alpha]$ is finitely generated
	by some $b_1, \dots, b_m$.
	Viewing the $b_i$ as polynomials in $\alpha$, we can select a large integer
	$N$ (say $N = \deg b_1 + \dots + \deg b_m + 2015$)
	and express $\alpha^N$ in the $b_i$'s to get
	\[ \alpha^N = c_1b_1(\alpha) + \dots + c_mb_m(\alpha). \]
	The above gives us a monic polynomial in $\alpha$,
	and the choice of $N$ guarantees it is not zero.
	So $\alpha$ is an algebraic integer.
\end{proof}
\begin{example}[$\half$ isn't an algebraic integer]
	We already know $\half$ isn't an algebraic integer.
	So we expect 
	\[
		\ZZ \left[ \half \right]
		= \left\{ \frac{a}{2^m} \mid a, m \in \ZZ \text{ and } m \ge 0 \right\}.
	\]
	to not be finitely generated, and this is the case.
\end{example}
\begin{ques}
	To make the last example concrete:
	name all the elements of $\ZZ[\half]$
	that cannot be written as an integer combination of
	\[ \left\{ \frac12, \frac{7}{8}, \frac{13}{64},
		\frac{2015}{4096}, \frac{1}{1048576} \right\} \]
\end{ques}

Now we can state the theorem.
\begin{theorem}[Algebraic integers are closed under $+$ and $\times$]
	The set $\ol\ZZ$ is closed under addition and multiplication;
	i.e.\ it is a ring.
	In particular, $\OO_K$ is also a ring for any number field $K$.
\end{theorem}
\begin{proof}
	Let $\alpha, \beta \in \ol\ZZ$.
	Then $\ZZ[\alpha]$ and $\ZZ[\beta]$ are finitely generated.
	Hence so is $\ZZ[\alpha, \beta]$.
	(Details: if $\ZZ[\alpha]$ has $\ZZ$-basis $a_1, \dots, a_m$ and
	$\ZZ[\beta]$ has $\ZZ$-basis $b_1, \dots, b_n$,
	then take the $mn$ elements $a_ib_j$.)

	Now $\ZZ[\alpha \pm \beta]$ and $\ZZ[\alpha \beta]$ are subsets of $\ZZ[\alpha,\beta]$ and so they are also finitely generated.
	Hence $\alpha \pm \beta$ and $\alpha\beta$ are algebraic integers.
\end{proof}

In fact, something even better is true.
As you saw, for $\QQ(\sqrt 3)$ we had $\OO_K = \ZZ[\sqrt 3]$;
in other words, $\OO_K$ was generated by $1$ and $\sqrt 3$.
Something similar was true for $\QQ(\sqrt 5)$.
We claim that in fact, the general picture looks exactly like this.

\begin{theorem}[$\OO_K$ is a free $\ZZ$-module of rank $n$]
	Let $K$ be a number field of degree $n$.
	Then $\OO_K$ is a free $\ZZ$-module of rank $n$,
	i.e.\ $\OO_K \cong \ZZ^{\oplus n}$ as an abelian group.
	In other words, $\OO_K$ has a $\ZZ$-basis of $n$ elements as
	\[ \OO_K
		= \left\{ c_1\alpha_1 + \dots
			+ c_{n-1}\alpha_{n-1} + c_n\alpha_n
			\mid c_i \in \ZZ \right\} \]
	where $\alpha_i$ are algebraic integers in $\OO_K$.
	\label{thm:OK_free_Z_module}
\end{theorem}
\begin{proof}
	I find this kind of fun: \Cref{prob:OK_free_Z_module}.
\end{proof}

This last theorem shows that in many ways $\OO_K$ is a ``lattice'' in $K$.
That is, for a number field $K$ we can find $\alpha_1$, \dots, $\alpha_n$
in $\OO_K$ such that
\begin{align*}
	\OO_K &\cong \alpha_1\ZZ \oplus \alpha_2\ZZ \oplus \dots \oplus \alpha_n\ZZ \\
	K &\cong \alpha_1\QQ \oplus \alpha_2\QQ \oplus \dots \oplus \alpha_n\QQ
\end{align*}
as abelian groups.

\section{Primitive element theorem, and monogenic extensions}
\prototype{$\QQ(\sqrt3,\sqrt5) \cong \QQ(\sqrt3+\sqrt5)$. Can you see why?}
I'm only putting this theorem here because I was upset that no one
told me it was true (it's a very natural conjecture),
and I hope to not do the same to the reader.
However, I'm not going to use it in anything that follows.

\begin{theorem}
	[Artin's primitive element theorem]
	Every number field $K$ is isomorphic to $\QQ(\alpha)$
	for some algebraic number $\alpha$.
	\label{thm:artin_primitive_elm}
\end{theorem}
The proof is left as \Cref{prob:artin_primitive_elm}, since to prove it I need to talk
about field extensions first.

The prototypical example \[ \QQ(\sqrt3,\sqrt5) \cong \QQ(\sqrt3+\sqrt5) \]
makes it clear why this theorem should not be too surprising.
%To see why this is true, note that $K$ contains the element
%\[ \left( \sqrt3+\sqrt5 \right)^2-8 = 2\sqrt15 \]
%and hence also the element $2(\sqrt15)(\sqrt3+\sqrt5) = 6\sqrt5+10\sqrt3$.
%Thus from the fact that $6\sqrt5+10\sqrt3$ and $\sqrt3+\sqrt5$ are in $K$,
%we can extract both $\sqrt 3$ and $\sqrt 5$.
%Thus $\sqrt3, \sqrt5 \in K$, which is what we wanted.

\bigskip

We might hope that something similar is true of the ring of integers:
that we can write \[ \OO_K = \ZZ[\theta] \]
in which case $\{1, \theta, \dots, \theta^{n-1}\}$
serves both as a basis of $K$ and as the $\ZZ$-basis for $\OO_K$ (here $n = [K:\QQ]$).
In other words, we hope that the basis of $\OO_K$is actually a ``power basis''.

This is true for the most common examples we use:
\begin{itemize}
	\ii the quadratic field, and
	\ii the cyclotomic field in \Cref{prob:ring_int_cyclotomic}.
\end{itemize}
Unfortunately, it is not true in general:
the first counterexample is $\QQ(\alpha)$ for $\alpha$ a root of $X^3-X^2-2X-8$.

We call an extension with this nice property \vocab{monogenic}.
As we'll later see, monogenic extensions have a really nice factoring algorithm,
\Cref{thm:factor_alg}.

\section\problemhead
\begin{sproblem}
	\label{prob:OK_unit_norm}
	Show that $\alpha$ is a unit of $\OO_K$ (meaning $\alpha\inv \in \OO_K$)
	if and only if $\Norm_{K/\QQ}(\alpha) = \pm 1$.
	\begin{hint}
		The norm is multiplicative and equal to product of Galois conjugates.
	\end{hint}
\end{sproblem}
% trivial

\begin{problem}
	[Black MOP 2010]
	There are $n > 2$ lamps arranged in a circle; initially one is on and the others are off.
	We may select any regular polygon whose vertices are among the lamps
	and toggle the states of all the lamps simultaneously.
	Show it is impossible to turn all lamps off.
	\begin{hint}
		View as roots of unity. Note $\half$ isn't an algebraic integer.
	\end{hint}
\end{problem}

\begin{problem}
	[USA TST 2012] Decide whether there exist $a,b,c > 2010$ satisfying
	\[ a^3+2b^3+4c^3=6abc+1. \]
	\begin{hint}
		Norm in $\QQ(\sqrt[3]2)$.
	\end{hint}
\end{problem}

\begin{problem}
	\label{prob:OK_free_Z_module}
	\gim
	Let $K$ be a number field of degree $n$.
	Using the norm, show that $\OO_K$ is a free $\ZZ$-module of rank $n$
	(thus proving \Cref{thm:OK_free_Z_module}).
	\begin{hint}
		Pick a $\QQ$-basis of $\alpha_1$, \dots, $\alpha_n$ of $K$ and WLOG
		the $\alpha_i$ are in $\OO_K$ by scaling.
		Look at $\Norm(c_1\alpha_1 + \dots + c_n\alpha_n)$.
		Bound denominators.
	\end{hint}
	\begin{sol}
		Pick a $\QQ$-basis of $\alpha_1$, \dots, $\alpha_n$ of $K$ and WLOG
		the $\alpha_i$ are in $\OO_K$ by scaling.
		Pick $\alpha \in \OO_K$
		Look at $\Norm(\alpha) = \Norm(c_1\alpha_1 + \dots + c_n\alpha_n)$.
		If we do a giant norm computation, we find that $\Norm(\alpha)$
		in the $c_i$ with fixed coefficients.
		(For example, $\Norm(c_1 + c_2\sqrt 2) = c_1^2 - 2c_2^2$, say.)
		But $\Norm(\alpha)$ is an \emph{integer}, so the denominators of the $c_i$ 
		have to be bounded by some very large integer $N$.
		Thus
		\[
			\bigoplus_i \ZZ \cdot \alpha_i
			\subseteq 
			\OO_K
			\subseteq
			\frac 1N \bigoplus_i \ZZ \cdot \alpha_i.
		\]
		The latter inclusion shows that $\OO_K$ is a subgroup
		of a free group, and hence it is itself free.
		On the other hand, the first inclusion shows it's rank $n$.
	\end{sol}
\end{problem}

\begin{dproblem}[Cyclotomic Field]
	\yod
	\label{prob:ring_int_cyclotomic}
	Let $p$ be an odd rational prime and $\zeta_p$ a primitive $p$th root of unity.
	Let $K = \QQ(\zeta_p)$.
	Prove that $\OO_K = \ZZ[\zeta_p]$.
	(In fact, the result is true even if $p$ is not a prime.)
	\begin{hint}
		Obviously $\ZZ[\zeta_p] \subseteq \OO_K$, so our goal is to show the reverse inclusion.
		Show that for any $\alpha \in \OO_K$, the trace of $\alpha(1-\zeta_p)$ is divisible by $p$.
		Given $x = a_0 + a_1\zeta_p + \dots + a_{p-2}\zeta^{p-2} \in \OO_K$ (where $a_i \in \QQ$),
		consider $(1-\zeta_p)x$.
		\todo{add solution from math 129 notes}
	\end{hint}
\end{dproblem}

\begin{problem}[Kronecker's theorem] \yod
	Let $x$ be an algebraic integer.
	Suppose all its conjugates have absolute value one.
	Prove that $x^N=1$ for some positive integer $N$.
\end{problem}

\begin{problem}
	\label{prob:rep_lemma}
	Consider $n$ roots of unity $\eps_1$, \dots, $\eps_n$.
	Assume the average $\frac1n(\eps_1 + \dots + \eps_n)$ is an algebraic integer.
	Prove that either the average is zero or $\eps_1 = \dots = \eps_n$.
	(This lemma will be used when we prove Burnside's theorem using representation theory.)
\end{problem}
